
\subsection{Project Evaluation}

The software project created as part of this thesis has been evaluated according to \cref{sec:method-open-source}.
The results of the evaluation are shown in \cref{tab:open-source-evaluation}.
Some results are clarified further below, using the test requirement's ID as the paragraph name.

\paragraph{OS8} The project has a Changelog.md file for the \gls{VSCode} extension, but not for the other components or the project as a whole.

\paragraph{OS10} There is no official plan. But the project could potentially be promoted using EclipseCon 2021, or through email to related stakeholders such as EclipseSource members (like Dr. Jonas Helming).

\paragraph{OS11} The author is the committed person, but the duration of commitment cannot be guaranteed to be long enough.

\paragraph{OS18} Not all the code is commented, but the (subjectively) required parts are.

\input{tables/evaluation-open-source-project.tex}

\subsection{Readme Evaluation}

The root level ``Readme'' file is very important.
It is the face of the project to the world.
Many of the \gls{open source} evaluation elements are related to the Readme.
Therefore, they have been condensed to a separate evaluation.
Note that this project has multiple modules, and therefore multiple Readme files.
This evaluation only looks at the top level, project wide Readme.
The results of the evaluation are shown in \cref{tab:readme-evaluation}.
Some results are clarified further below, using the test requirement's ID as the paragraph name.

\paragraph{OSR3} The Readme does not have the description, but the \gls{GitHub} project has a description shown in the side of the webpage of the project.

\paragraph{OSR5} The Readme does not have a project background story, but the \gls{GitHub} project has a description explaining this is a result of a masters thesis.

\input{tables/readme-evaluation.tex}
