\chapter*{Abstract}

Software engineering has an approach called Model-Driven Development (MDD).
This is taught to students in higher education.
The approach is reliant on tools, and one such tool is the Eclipse Modeling Framework (EMF).
While EMF can be used to teach students about MDD, it is unpopular because of its ties to the Eclipse Integrated Development Environment (IDE), causing students to resist learning MDD.
Cloud based alternatives exist for the Eclipse IDE, like Gitpod with VSCode, which provide benefits in an educational organization.
However, the EMF tools used in education are not available in these alternatives.
This thesis tries to enable the cloud based alternatives to support EMF.

The thesis' approach is based on Design Science Research, where a design is created and a software artifact is implemented.
The design draws inspiration from the Language Server Protocol (LSP) and Graphical Language Server Platform (GLSP), protocols for text and diagram editing.
These protocols already work in VSCode.

The result is a Tree Editor extension for VSCode.
EMF models can be edited as trees.
This extension uses a three component design: a generic tree editor user interface, a VSCode extension, and an EMF specific server.
The extension and server communicate with a newly designed protocol, the \acrfull{TLSP}.

The resulting implementation can be built upon to work with \acrshort{EMF} modeling in the cloud.
The \acrshort{TLSP} protocol and software architecture can be used by other tools that need tree editors, that aim to support multiple IDEs as well.
A wider adoption of \acrshort{TLSP} in IDEs will make migrations of tree editors to other IDEs easy.
Regardless, the design provides a reusable server for EMF, which can ease migrations of EMF to other IDEs.
