\chapter{Results}\label{chap:results}

This chapter will present the outcomes of the design and development phase.
First, the artifact is presented from a user's point of view, showcasing functionality with screenshots.
Then, the artifact's design will be presented from an architectural view, and the protocol will be presented last.


The results will also present the measures taken for making the \gls{open source} project viable, and something a developer community may want to develop and maintain further.

Evaluation of the results is presented later, in \cref{chap:evaluation}.

\section{Software Artifact: Tree Editor Extension for Ecore in Gitpod}

This is of interest for a stakeholder, and someone aiming to do further research on this design.

\input{chapters/6-results/0-tree-editor.tex}

* Functioning viewer of any Ecore model in VSCode. Missing most features.
  * List of functional requirements implemented.
  * List of non-functional requirements implemented.
* Reuses java EMF code and EMF.Cloud ModelServer. No re-implementation of EMF logic.

* TLSP architecture. 3-component system (view, extension, server).


\section{Design Artifact: Architecture for Tree Language Server Systems}

\input{chapters/6-results/1-architecture.tex}


\section{Design Artifact: Tree Language Server Protocol}\label{sec:tlsp}
% TODO

\input{chapters/6-results/2-protocol.tex}

* TLSP protocol. JSON-RPC based with LSP Base, with a Tree Document Model data structure. Hierarchy schemas for optimistic-view drag-n-drop. Type and id based nodes. 


\section{Open Source Project: Measures Taken for Viability and Maintainability}
% TODO

* A project that may be viable and further developed.
  * Bundled IDE configuration: recommended extension, build tasks, run configurations. Reduces overhead for new developers.
  * NOT: CI/CD. Easily added later when needed. Overhead for 1-man project.
  * Build-scripts and npm/maven for building.
  * Readme-files for all components and modules.
  * Commonly used programming languages and (official) dependencies
  
