\chapter{Discussion}\label{chap:discussion}

\input{chapters/8-discussion/1-discussion.tex}


\section{Limitations}

Research is not always perfect, and creative design is hard to do in an objective manner.
This section addresses some limitations, shortcomings and uncertainties of this thesis.

\paragraph{An unufinished instantiation leaves risk in the design}
Because development takes longer time than what is allocated for a master's thesis, a substantial amount the protocol design is left on a theoretical design stage.
Many challenges and constrains appear first when the design is implemented, imposed by the surrounding frameworks, \acrshortpl{API} and tools used to create the software.
Although the author is confident in this thesis' use case evaluation, the results cannot be guaranteed with the utmost certainty before they are all implemented.
Regardless, the overarching design and protocol should still be valid and contribute new knowledge.
Most obstacles encountered from here on, are assumed to be possible to work around with minor changes to the protocol.

\paragraph{Agile development causes a protocol to emerge slowly}
Doing an Agile approach means creating software that works.
It also means postponing work, instead of having a lot of unfinished work in progress.
This is good for delivering working software to customers and users, but not for developing a full and complete protocol specification.
While an agile approach can always back up the validity and usefulness of protocol elements with a working implementation, it may leave unidentified risks for later, contrary to how research aims to tackle risk early.
It can be argued, however, that Design Science Research uses the build-evaluate cycle exactly to produce correct knowledge, instead of drafts and unproven hypotheses.

\paragraph{Lacking contact with stakeholders and Eclipse ecosystem}
The developers of tools like Theia Tree Editor, ecore-glsp, JSON-Forms etc. may have deeper insights and knowledge than the author.
They may be able to suggest better, more pragmatic and effective design solutions.
They can also verify or disprove, or at least indicate, the need for generic tree editors and protocols for \acrshort{EMF}.
The author did not actively communicate with the Eclipse ecosystem, outside of two private chats during EclipseCon 2020.
There has also not been contact with other students of \gls{TDT4250}.
The main reason is that the course is taught in the autumn, while the master thesis is done during the spring semester.
However, the author has been a student of \gls{TDT4250} and talked to other classmates during the semester for their opinion on \acrshort{EMF} and \gls{Eclipse}.

\paragraph{Design Science Research methodologies can cause ad-hoc evaluations}
The literature on Design Science Research says little on how to evaluate.
There are only abstract, high level recommendations to the forms of evaluation, such as action research, case study, simulation and so on.
This can cause a researcher to create an evaluation which is ultimately positive of their view, or disregards important aspects.
The author has attempted to keep the evaluations related to the original research goals, and grounded in existing practice.
Regardless, the weakness of unguided evaluation design is there.


\section{Further Work}

This thesis has uncovered some new potential areas of research.

\paragraph{Theia or VSCode as a deployment platform}
The GenModel and code generator can target \gls{Eclipse} as a deployment platform.
The model gets a plugin generated, and an editor in \gls{Eclipse} for model instances.
A similar approach can be interesting for deploying a tree editor for model instances, but using \gls{Theia} or \gls{VSCode} instead of \gls{Eclipse}.
These could potentially reuse the \acrshort{TLSP} as well.
Doing this would increase the value of \acrshort{EMF}, and further illustrate the values of \acrlong{MDD} and code generation.

\paragraph{Collaborative modeling}
Modeling is a collaborative task.
With the move to cloud, and with the increased amount of work from home\footnote{Due to covid-19.}, collaboration can move online as well.
This is already normal for things like Google Docs, and Jetbrains just added ``Code With Me''.
Obeo is also developing this for their cloud based Sirius modeling tool.
When all the editing is done through a protocol like \acrshort{TLSP}, the data could be redirected to multiple clients, meaning multiple students' computers.

\paragraph{Completing the EMF editor}
A lot of the remaining work is routine design, not research.
But some remaining parts may be more challenging, such as getting the GenModel working and specializing the editor to properly show a \texttt{.genmodel} file like in \gls{Eclipse}.
There are also challenges to loading the generated code back into the server, for validations and custom \texttt{ItemProviders}.
Completing this will make the editor more useful, and also move it closer to a solution suited for industry use, beyond just education.


